---
title: "Modifying biproporz()"
date: 2025-03-05
author: "Flavio Poletti"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying biproporz()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.kable.NA = "")
```

## Introduction

This vignette demonstrates how to modify and extend the default implementation of biproportional apportionment in `biproporz()`. We cover the following modifications:

- Applying a quorum to filter seat-eligible parties
- Using alternative predefined apportionment methods
- Implementing a custom rounding function for the lower apportionment
- Altering the Winner-Take-One (WTO) method to bypass upper apportionment constraints

We'll begin by creating a custom dataset: a matrix containing vote counts and a vector representing the number of seats per district.

```{r setup}
library(proporz)

# Define a custom dataset for this vignette
votes_matrix = matrix(
    c( 800, 2802, 4095,  0, 150,
      3900,  814, 3990, 20,  60,
      1400, 1302, 4305, 10,  80,
         0,    0,    0, 50,   0,
       610,  500, 1001, 40, 120),
    ncol = 5, byrow = TRUE,
    dimnames = list(
        party = c("A", "B", "C", "D", "E"),
        district = c("City 1", "City 2", "City 3", "Region 4", "Region 5")
    ))

district_seats = setNames(c(5, 5, 14, 1, 1), colnames(votes_matrix))
```

### Analyzing the Votes Matrix

In our example `votes_matrix`, each voter casts as many votes as there are seats in the district.
To understand the party strength across districts, we calculate the number of *voters* per party and district by weighting the votes matrix with `weight_list_votes()`:

```{r weight_votes}
votes_matrix

(voters = weight_list_votes(votes_matrix, district_seats))
```

This step is performed within `biproporz()` with the parameter `use_list_votes = TRUE`.
The result is a matrix with fractional voter counts, as votes can be split between multiple parties.

## Standard Biproportional Apportionment

The default `biproporz()` method allocates seats based on the votes matrix and district seats using standard rounding for both the upper and lower apportionments.

```{r standard}
(v1 = biproporz(votes_matrix, district_seats))

# Number of seats per party
rowSums(v1)
```

Key observations from the standard apportionment:

- *Party D* did not receive enough votes across all districts to win a seat.
- In *City 3*, *Party B* receives the most seats despite having fewer total votes than two other parties.
- *Party E*'s strongest district is *City 1* (122 voters), but it does not win a seat there.

The `biproporz()` function returns the seat allocation as a matrix with divisors as hidden attributes (see `get_divisors()`).
You can print the marginal sums (total seats per party and district) and divisors with `summary()`.

```{r summary}
summary(v1)

# You can transpose the matrix
# summary(t(v1))
```

## Modifying biproporz with Parameters

### Applying a Quorum

A quorum can ensure that only parties with a minimum percentage or number of votes are eligible for seat allocation.
In this example, we impose a 5% quorum on the total votes.

```{r standard_quorum}
(v2 = biproporz(votes_matrix, district_seats, quorum_any(total = 0.05)))
```

This does not actually change the seat distribution compared to the standard method, as *Party D* is already below the "natural quorum" and did not get enough votes for any seat.

### Alternative Apportionment Methods

`biproporz()` allows you to specify different methods for the upper and lower apportionment by passing a list to the `method` parameter.
For example, we can use the Adams method for the upper apportionment and standard rounding for the lower apportionment.

```{r alternative_methods}
(v3 = biproporz(votes_matrix, district_seats, method = list("adams", "round")))
```

### Custom Rounding Function

To customize seat allocation, you can define your own rounding function for the lower apportionment.
Here’s a custom rounding function that works as follows:

- Values below 0.7 are rounded down to 0
- Values equal to or greater than 0.7 are rounded up to 1
- Standard rounding applies for values greater than 1 (i.e. values below .5 are rounded down)

```{r custom_rounding}
custom_rounding_func = function(x) {
    stopifnot(all(x >= 0))
    lt0.7 = x < 0.7
    x[lt0.7] <- 0
    x[!lt0.7] <- ceil_at(x[!lt0.7], 0.5)
    x
}

# The function must work with a matrix
custom_rounding_func(matrix(c(0.5, 0.6, 1.5, 2.5), 2))

# Apply custom rounding function in lower apportionment
(v4 = biproporz(votes_matrix, district_seats, 
                method = list("adams", custom_rounding_func)))
```

Compared to using standard rounding, parties *A* and *E* swap one seat in cities *1* and *2*.

## District Winner Methods

### Winner-Take-One (WTO)

The WTO method guarantees that the party with the most votes in each district will receive at least one seat, given that the party is eligible for a seat from the upper apportionment.
These two constraints may lead to conflicts and an error, as seen below.

```{r wto}
try(biproporz(votes_matrix, district_seats, method = "wto"))
```

*Party D* has the most votes in *Region 4* and should get a seat there but there's no party seat to allocate.
To prevent this case, a quorum is usually applied to ensure that only large enough parties are eligible for seats.

### WTO with an Alternative Upper Apportionment Method

By switching to the Adams method for the upper apportionment, *Party D* gets one seat (as we've already seen).
This resolves the conflict, allowing WTO to work in the lower apportionment.

```{r wto_other_method}
(v5 = biproporz(votes_matrix, district_seats, method = list("adams", "wto")))
```

### Guaranteeing District Winners a Seat

You can modify the WTO method to ensure district winners always get a seat, even if they don't meet the upper apportionment criteria.
Below is a custom function that implements this approach.
This is a non-standard approach and the function should be adapted as needed.

```{r absolute_wto_function}
biproporz_absolute_wto = function(votes_matrix, district_seats,
                                  quorum = NULL, use_list_votes = TRUE) {
    # 1) Identify unambiguous district winners
    # Note: This step could also happen after the quorum has been applied
    # (depending on the desired method implementation)
    district_winners = district_winner_matrix(votes_matrix, district_seats)
    district_winners[is.na(district_winners)] <- FALSE  # Ignore ties

    # 2) Apply quorum if specified
    if(!is.null(quorum)) {
        votes_matrix <- apply_quorum(votes_matrix, quorum)
    }
    
    # 3) Assign party seats in upper apportionment
    ua = upper_apportionment(votes_matrix, district_seats, 
                             use_list_votes, method = "round")

    # 4.1) Assign seats to district winners without 
    # enough upper apportionment seats
    seats_without_ua = district_winners * 1
    seats_without_ua[rowSums(district_winners) <= ua$party, ] <- 0
    
    # 4.2) Biproportional apportionment for remaining seats
    # Build votes matrix, set votes for district winners 
    # without enough upper apportionment seats to zero
    biprop_votes_matrix = votes_matrix
    biprop_votes_matrix[seats_without_ua > 0] <- 0
    
    # Reduce the number of seats for districts that 
    # already had a "insufficient district winner" seat assigned
    non_biprop_distr = colSums(seats_without_ua) > 0
    biprop_district_seats = district_seats
    biprop_district_seats[non_biprop_distr] <- 
        biprop_district_seats[non_biprop_distr] - 1
    
    # Run biproporz
    seats_biproporz = biproporz(biprop_votes_matrix, biprop_district_seats, 
                                method = "wto")
    
    # Remove divisor attributes, as they're no longer 
    # meaningful for the combined distribution
    seats_biproporz <- as.matrix(seats_biproporz)
    
    # 5) Return final seat distribution,
    #    combining the two apportionments 
    return(seats_biproporz + seats_without_ua)
}
```

Let’s compare the standard biproportional apportionment to the modified method, which guarantees district winners a seat.

```{r absolute_wto}
(v1 = biproporz(votes_matrix, district_seats)) # same as above

(v6 = biproporz_absolute_wto(votes_matrix, district_seats))

# Show the difference
v6 - v1
```

In this example, *Party D* gains a seat from *Party B*. To satisfy the new constraints, there are changes in district seat distributions for *Party A* and *Party E*.

Note that while this method resolves possible conflicts for district winners without upper apportionment seats, other conflicts might arise.
For example, if too many district winners are missing upper apportionment seats, the constraints for allocation the remaining seats may become overly restrictive.
